Panthera: Caching and Cache-based Scheduling in Distributed Computing Systems

### 2014-05-13 paper/paper.tex

\section{Introduction}
As computational problems and their associated datasets have grown in size, it has become necessary to take advantage of distributed computing systems to solve them \cite{taylor2010overview}. Such systems allow programs to scale from one computer (i.e. one \textit{node}) to thousands of computers with few or no changes in the codebase. In particular, the Hadoop distributed system \cite{hadoop} has seen tremendous growth. Hadoop is an open source version of the revolutionary MapReduce system developed at Google \cite{mapreduce}. With it, developers can easily take advantage of large, multi-node clusters to solve computational problems.

Current Hadoop deployments exist in the thousands of nodes at companies such as Cloudera, Facebook, Yahoo, etc. Especially in such large networks, \textit{access latency}, or the waiting time associated with accessing files through the Hadoop Distributed File System (\textit{HDFS}), is a significant concern. Latency in distributed systems can have significant effects and a reduction of the same comes with tremendous benefits. As the RAMCloud project has outlined \cite{ramcloud}, low latency can greatly extend the applications of distributed computing systems. For example, tree traversal algorithms are largely impractical in high latency systems such as Hadoop \cite{ramcloud}, however, with lowered latency, such algorithms may become useful and applicable. HDFS currently does not use local (i.e. on the client) random access memory (RAM) to improve latency.

In this paper, I present \textit{Panthera}, a \textit{cache layer} and \textit{cache-based scheduling system} for Hadoop. The \textit{Panthera} cache layer functions by retaining recently accessed files in random access memory, thus, when the files are needed once again, they do not have to be downloaded or read from the hard drive, thereby greatly reducing file access latency. Another portion of the project, the \textit{Panthera} scheduling system utilizes information about the data blocks cached in a cluster at a given time to schedule jobs. By doing so, it aims to maximize the benefit of caching to a series of jobs.

In this paper, we discuss both the \textit{Panthera} caching layer and scheduling system. In Section 2, we discuss the specific challenges associated with such a project. Section 3 details the architecture differences between a standard Hadoop installation and one with \textit{Panthera}.  Sections 4 and 5 detail the differences between metadata and data caching within \textit{Panthera}. Results and conclusions are covered in Sections 6, 7, 8, 9, concluding with a roadmap for future development in Section 10.

\section{Design Constraints}

There are several systems built on Hadoop that are in widespread use \cite{hbase, cloudbatch, pig}. The Hadoop project is also supported by large corporations with a 
consistent release cycle. Thus, for \textit{Panthera} to be practical, it must operate independently of the existing Hadoop codebase, i.e. as a layer, rather than modification. This implies that \textit{Panthera} must implemented without a single change in the existing Hadoop codebase, i.e. as a "drop-in" solution. Additionally, for non-cache related requests, \textit{Panthera} must add insignificant latency. Finally, data and metadata request latency should be significantly lowered with \textit{Panthera} in comparison to a standard Hadoop installation.

In order to develop a cache-based scheduling system, the \textit{Panthera} caching layer must be able to report information (e.g. blocks currently cached) over the network. The cache-based scheduler then uses information about blocks cached and the files each job requires to run in order to optimize the running order of a set of jobs (see Section \ref{scheduler-section} for a complete description of the \textit{Panthera} scheduling algorithm).

\section{Metadata and Data}

Within the Hadoop File System (HDFS), there is a clear distinction between \textit{metadata} and \textit{data} \cite{hdfs}. Metadata refers to information about a file or a directory, e.g. file size, file names, on which server the contents of a file are located, etc. For HDFS, the NameNode server handles all metadata. For every request, the NameNode recomputes metadata. \textit{Panthera} can reduce this computation by caching the results of a previous request, thereby conserving resources.

@@

\section{Architecture}
\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.4]{assets/hadoop_architecture.pdf}
	\includegraphics[scale=0.4]{assets/panthera_architecture.pdf}
  \caption{Standard Hadoop Architecture (left), \textit{Panthera}-Hadoop Architecture (right)}
  \label{arch}

\end{figure}

Figure~\ref{arch} outlines the architecture differences between a standard Hadoop File System architecture and an installation with \textit{Panthera}. In HDFS, the \textit{NameNode} serves as a metadata server, handling all requests from clients relating to metadata (e.g. directory listing, file sizes, block locations, etc.). \textit{DataNode(s)} hold data and serve it at request from the client. A more complete description of the Hadoop File System may be found in \cite{hadoop}.

\textit{Panthera} operates by intercepting requests and responses at the NameNode and the client. Both nodes have a running instance of \textit{Panthera}. Every request issued from the client is first reviewed by \textit{Panthera}, and if possible, answered immediately from the cache. The \textit{Panthera} instance on the NameNode constantly monitors the filesystem for changes that need to be reflected in the cache (e.g. if a file is deleted, the cache on the client should not have a copy of it).

\textit{Panthera}, since it caches both data and metadata, can be considered two separate subsystems, which are now discussed.

\section{Metadata caching}

@@

The \textit{Panthera} metadata system runs on the NameNode and communicates with the \textit{Panthera} instance on the client. The challenge on the metadata portion consist of reducing computational latency to memory latency. For example, for some intensive methods (e.g. a recursive directory listing), it may be worthwhile to cache a response in order to prevent computing the same list again.

However, the problem lies in identifying such methods. Obviously, there are some methods for which caching is not worthwhile since their execution time is lower or comparable to memory latency. Within \textit{Panthera}, six different Hadoop query methods were identified as medium to high latency. \textit{Panthera} identifies packets that contain these methods and subsequently caches them.

Currently, a standard LRU cache \cite{cache} is used for each method. Testing of different caching methods and predictive prefetching algorithms is under way.

\section{Data caching}

@@

The challenges with data caching differ significantly from those in metadata caching. Here, the goal is to convert network and hard drive read latency into memory latency. As \cite{ramcloud, hadoop} note, memory latency can be extremely low in modern systems, especially in comparison to network and hard drive latencies. Thus, in some respects, the goal of lowered latency is "easier" to attain in the data cache. However, for smaller files, it may be difficult to obtain latency improvement.

Hadoop, in professional and academic environments, is often used with series rather than individual jobs. Deciding how to order these jobs depending on cached file availability is a problem of great interest; \textit{Panthera} presents the solution of a cache-based scheduler for Hadoop.

\section{Scheduling Algorithm} \label{scheduler-section}

@@

\section{Testing Methodology}

In order to test the metadata cache, a directory listing query was repeatedly run on a directory with 100 files in it, and the latency times for responding are measured, with \textit{Panthera} and without. The reasoning for picking such a test lies in the fact that a directory listing in Hadoop involves all six of the cached methods as well as a few non-cached methods. Thus, the test provides a good estimate of the efficacy of the cache. Latency times for the data cache were obtained by repeatedly querying for a 64 megabyte large file. The file size was chosen since it is the default block size for a Hadoop cluster and the \textit{Panthera} cache stores files as blocks. Both metadata and data cache testing was conducted on virtualized server instances running on hardware with solid state hard drives.

@@

\section{Results}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{assets/v2/getter_boxplot.pdf}
  \caption{Data access latency}
  \label{data}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{assets/v2/multinode_meta_box_plot.pdf}
  \caption{RefCount job performance}
  \label{meta}
\end{subfigure}
\caption{Metadata access latency}
\label{fig:test}
\end{figure}

@@

\section{Conclusion}

@@

\section{Applications}
One of the major benefits of the systems developed is that they are independent of the existing Hadoop codebase and they do not require modifications to existing Hadoop applications. Thus, \textit{Panthera} may be deployed in a wide range of scenarios, such as the following:

@@

\section{Future work}
One of the most immediate avenues for further research is in \textit{domain-specific prefetching}. Different problem domains (e.g. natural language processing, predictive analytics, etc.) have different dataset requirements. It may be possible to predict what files a given job will need depending
on what problem domain the job is in. A generalized version of the same problem would be \textit{predictive prefetching} \cite{griffioen1994reducing} which involves the prediction and prefetching of files that will be required in the future. In both these problems, the Hadoop environment provides unique
challenges.

>>>>

\section{Introduction}
As computational problems and their associated datasets have grown in size, it has become necessary to take advantage of distributed computing systems to solve them \cite{taylor2010overview}. Such systems allow programs to scale from one computer (i.e. one \textit{node}) to thousands of computers with few or no changes in the codebase. In particular, the Hadoop distributed system \cite{hadoop} has seen tremendous growth. Hadoop is an open source implementation of the revolutionary MapReduce system developed at Google \cite{mapreduce}. With it, developers can easily take advantage of large, multi-node clusters to solve computational problems.

Hadoop consists of a computational layer termed \textit{MapReduce} and a custom distributed filesystem named \textit{Hadoop Distributed File System (HDFS)}. HDFS is used to allow all nodes in a cluster access to the same dataset. For instance, one may consider running an algorithm that counts the frequency of each word in a large text file. If this algorithm is run on a Hadoop cluster, each node must have access to the file in question in order to determine word frequency.

Current Hadoop deployments exist in the thousands of nodes at companies such as Cloudera, Facebook, Yahoo, etc. Especially in such large networks, \textit{access latency}, or the waiting time associated with accessing files and related information through the Hadoop Distributed File System (\textit{HDFS}), is a significant concern \cite{hdfs}. Latency in distributed systems can have significant effects and a reduction of the same comes with tremendous benefits. As the RAMCloud project has outlined \cite{ramcloud}, low latency can greatly extend the applications of distributed computing systems. For example, algorithms that require files other than the input file are largely impractical in high latency systems such as Hadoop \cite{ramcloud, spark}. A similar case exists with jobs (or sets of jobs) that consist of multiple passes over the same dataset; many of these exist in the field of machine learning, analytics, etc. \cite{spark}. However, with lowered latency, such algorithms may become useful and applicable. HDFS currently does not use local (i.e. on the client) random access memory (RAM) to improve latency.

I present \textit{Panthera}, a \textit{cache layer} and \textit{cache-based scheduling system} for Hadoop. The \textit{Panthera} cache layer functions by retaining recently accessed files and related information in random access memory. Thus, when the files are needed once again, they do not have to be downloaded or read from the hard drive, thereby greatly reducing file access latency. Another portion of the project, the \textit{Panthera} scheduling system utilizes information about the data blocks cached in a cluster at a given time to schedule jobs. By doing so, it aims to maximize the benefit of caching to a series of jobs.

In this paper, the \textit{Panthera} caching layer and scheduling system are presented. In Section \ref{sect:architecture}, motivation and construction of the \textit{Panthera} architecture are explained. Section~\ref{sect:caching_layer} describes the caching layer portion of this project. Section~\ref{sect:scheduler} presents the scheduling algorithm and implementation details. The results and data are noted in Section~\ref{sect:results}. Finally, Sections \ref{sect:conclusions}, \ref{sect:applications}, \ref{sect:future_work} conclude with an interpretation of the results, discussion of potential applications and future work.

\section{Architecture} \label{sect:architecture}
\subsection{Design Constraints}  \label{design_constraints}
The primary goal of the \textit{Panthera} project is to reduce access latency in HDFS. Therefore, data and metadata request latency should be significantly lowered with \textit{Panthera} in comparison to the control group of a standard Hadoop installation.

There are many systems built on Hadoop that are in widespread use \cite{hbase, cloudbatch, pig}. The Hadoop project is also supported by large corporations with a 
consistent release cycle. Thus, for \textit{Panthera} to be practical, it must opedrate independently of the existing Hadoop codebase, i.e. as a layer, rather than modification. This implies that \textit{Panthera} must implemented without a single change in the existing Hadoop codebase, i.e. as a ``drop-in'' solution. This is due to the fact that a deviation from the mainline Hadoop codebase would have to updated at the same pace as the vanilla Hadoop, which would be contrary the benefits of Hadoop's large and active development community.  Also, since \textit{Panthera} does not require any code modifications to existing Hadoop applications, allowing it to be easily integrated into current software stacks.

In order to develop a cache-based scheduling system, the \textit{Panthera} caching layer must be able to report information (e.g. blocks currently cached) over the network. The cache-based scheduler then uses information about blocks cached and the files each job requires to run in order to optimize the running order of a set of jobs (see Section \ref{sect:scheduling} for a complete description of the \textit{Panthera} scheduling algorithm).

\subsection{Metadata and Data} \label{metadata_and_data}

Within the Hadoop File System (HDFS), there is a clear architectural distinction between \textit{metadata} and \textit{data} \cite{hdfs}. Metadata refers to information about a file or a directory, e.g. file size, file names, the location of a file in the cluster, etc. For HDFS, the NameNode server handles all metadata. For every request, the NameNode recomputes metadata. \textit{Panthera} can reduce this computation by caching the results of a previous request, thereby conserving resources.

@@

\subsection{Panthera Architecture} \label{architecture_sect}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale=0.4]{assets/hadoop_architecture.pdf}
  \caption{Vanilla Hadoop architecture}
  \label{fig:vanilla_arch}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale=0.4]{assets/panthera_architecture.pdf}
  \caption{Panthera-Hadoop architecture}
  \label{fig:panthera_arch}
\end{subfigure}
\label{fig:arch}
\end{figure}

Figure~\ref{architecture_sect} outlines the architecture differences between a standard Hadoop File System architecture and an installation with \textit{Panthera}. In HDFS, the \textit{NameNode} serves as a metadata server, handling all requests from clients relating to metadata (e.g. directory listing, file sizes, block locations, etc.). \textit{DataNode(s)} hold data and serve it at request from the client. In order to read or write a file, a client first obtains permission from the NameNode and subsequently communicates with a DataNode to complete the operation. Though the aforementioned background is sufficient for this paper, a more complete description of the Hadoop File System may be found in \cite{hadoop}.

\textit{Panthera} operates by intercepting requests and responses at the NameNode and the client. Both nodes have a running instance of \textit{Panthera}. Every request issued from the client is first reviewed by \textit{Panthera}, and if possible, answered immediately from the cache. If the response is not found in the cache, certain requests are cached and subsequent responses are associated with the request in consideration.

The \textit{Panthera} instance on the NameNode serves as a cache refresh mechanism. Though sparse in the Hadoop environment since input files are usually not written to in the MapReduce computational paradigm, files that are cached may become out-of-date, i.e. the cache becomes "stale". The \textit{Panthera} instance on the NameNode monitors the filesystem for changes and reports certain changes to Panthera caches, which then remove stale files from the cache. The refresh mechanism is particularly important for metadata caching where the cached information is far more likely to change during the execution of a job.

The \textit{Panthera} cache layer, since it caches both data and metadata, can be considered two separate subsystems, which are now discussed.

\section{Panthera Caching Layer} \label{sect:caching_layer}
A significant portion of Panthera is a caching layer for Hadoop, the details of which are now covered. 
\subsection{Metadata caching} \label{sect:metadata_caching}

@@

The \textit{Panthera} metadata system runs on the NameNode and communicates with the \textit{Panthera} instance on the client. The challenge on the metadata portion consists of converting computational latency to memory latency. For example, for some intensive methods (e.g. a recursive directory listing), it may be worthwhile to cache a response in order to prevent computing the same list again.

However, the problem lies in identifying such methods. Obviously, there are some methods for which caching is not worthwhile since their execution time is lower or comparable to memory latency. Within Hadoop, there are several methods that incur very significant computational latency, e.g. a recursive directory listing. Only these methods are cached within Panthera.

Currently, a standard LRU cache \cite{cache} is used for each method. Testing of different caching methods and predictive prefetching algorithms \cite{griffioen} is under way.

\subsection{Data caching} \label{sect:data_caching}

@@

The challenges with data caching differ significantly from those in metadata caching. Here, the goal is to convert network and hard drive read latency into memory latency. As \cite{ramcloud, hadoop} note, memory latency can be extremely low in modern systems, especially in comparison to network and hard drive latencies. Thus, in some respects, the goal of lowered latency is ``easier'' to attain in the data cache in comparison to the metadata cache.

\section{Scheduler} \label{sect:scheduler}
\subsection{Hadoop Job Structure}
Hadoop, in professional and academic environments, is often used with series rather than individual jobs. For example, in machine learning, various algorithms (termed \textit{iterative machine learning algorithms}) consist of multiple Hadoop jobs rather than a single job \cite{spark}. In Hadoop-based querying systems \cite{hive, pig}, a set of queries may consist of jobs that are entirely independent of one-another. Deciding how to order these jobs depending on cached file availability is a problem of great interest; \textit{Panthera} presents the solution of a cache-based scheduler for Hadoop.

\subsection{Cluster Saturation}
It may be asked why it is not possible to run independent jobs in parallel on the Hadoop cluster. Each node in a given cluster has a maximum number of \textit{map} or \textit{reduce} calls (i.e. the basic computational units in Hadoop) it may process at a given time. If a job saturates nodes in a cluster, running another job concurrently would provide no running time benefit. Thus, a scheduler is useful since it may organize independent jobs in a beneficial ordering. In \textit{Panthera}, this ordering is based on the files or blocks present in the cluster's caches.

\subsection{Scheduling Algorithm} \label{sect:scheduling}

@@

\section{Results} \label{sect:results}

\subsection{Testing Methodology}
In order to test the metadata cache, a directory listing query was repeatedly run on a directory with 100 files in it, and the latency times for responding are measured, with \textit{Panthera} and without. The reasoning for picking such a test lies in the fact that a directory listing in Hadoop involves all six of the cached methods as well as a few non-cached methods. Thus, the test provides a good estimate of the efficacy of the cache. Latency times for the data cache were obtained by repeatedly querying for a 64 megabyte large file. The file size was chosen since it is the default block size for a Hadoop cluster and the \textit{Panthera} cache stores files as blocks. Both metadata and data cache testing was conducted on virtualized server instances running on hardware with solid state hard drives.

@@

\subsection{Results}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{assets/v2/getter_boxplot.pdf}
  \caption{Data access latency}
  \label{data}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{assets/v2/multinode_meta_box_plot.pdf}
  \caption{RefCount job performance}
  \label{meta}
\end{subfigure}
\caption{Metadata access latency}
\label{fig:test}
\end{figure}

@@

\section{Conclusions}\label{sect:conclusions}

@@

\section{Applications}\label{sect:applications}
One of the major benefits of the systems developed is that they are independent of the existing Hadoop codebase and they do not require modifications to existing Hadoop applications. Thus, \textit{Panthera} may be deployed in a wide range of scenarios, such as the following:

@@

\section{Future work}\label{sect:future_work}
One of the most immediate avenues for further research is in \textit{domain-specific prefetching}. Different problem domains (e.g. natural language processing, predictive analytics, etc.) have different dataset requirements. It may be possible to predict what files a given job will need depending
on what problem domain the job is in. A generalized version of the same problem would be \textit{predictive prefetching} \cite{griffioen1994reducing} which involves the prediction and prefetching of files that will be required in the future. In both these problems, the Hadoop environment provides unique
challenges.
